---
title: "Class 10: Halloween Project"
author: "Henry(A16354124)"
format: pdf
---

## Read document
```{r}
candy <- read.csv("https://raw.githubusercontent.com/fivethirtyeight/data/master/candy-power-ranking/candy-data.csv", row.names = 1)
head(candy)
```

```{r}
flextable::flextable(head(candy))
```

```{r}
library(dplyr)
candy %>% select(winpercent)
candy |> select(winpercent)
```


## Questions

> Q1. How many different candy types are in this dataset?

```{r}
nrow(candy)
candy |> nrow()
```

> Q2. How many fruity candy types are in the dataset?

```{r}
nrow(subset(candy, candy$fruity == 1))
```

> Q3. What is your favorite candy in the dataset and what is it’s winpercent value?

```{r}
# Definitely Eminem (M&M's)
candy["M&M's", "winpercent"]
```

> Q4. What is the winpercent value for “Kit Kat”?

```{r}
candy["Kit Kat", "winpercent"]
```

> Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy["Tootsie Roll Snack Bars", "winpercent"]
```

## The package "skimr"

```{r}
library("skimr")
skim(candy)
```

> Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

Yes, the variable "winpercent" is on a different scale compared to the other columns.

> Q7. What do you think a zero and one represent for the candy$chocolate column?

A zero represents that the candy does not contain chocolate, while a one indicates that the candy does contain chocolate.

> Q8. Plot a histogram of winpercent values

```{r}
library(ggplot2)

ggplot(candy) +
  aes(winpercent) +
  geom_histogram(binwidth = 5, fill = "lightblue", color = "black") +
  labs(title = "Histogram of Winpercent Values", x = "Winpercent", y = "Frequency") +
  theme_minimal(base_size = 12)
```

> Q9. Is the distribution of winpercent values symmetrical?

```{r}
ggplot(candy) +
  aes(winpercent) +
  geom_density()
```


No, the distribution of winpercent values is not symmetrical; it appears to be left-skewed.

> Q10. Is the center of the distribution above or below 50%?

```{r}
median(candy$winpercent) >= 50
mean(candy$winpercent) >= 50
summary(candy$winpercent)
```

> Q11. On average is chocolate candy higher or lower ranked than fruit candy?

```{r}
mean(candy$winpercent[candy$chocolate == 1]) > mean(candy$winpercent[candy$fruity == 1])
# 1.  Find all chocolate candy in the dataset
# 2. Extract their `winpercent` values
# 3. Find the mean of these values

# 4-6: Do the same for fruity candy
# 7. Which mean value is higher

mean_choc <- candy |> filter(chocolate == 1) |> summarise(mean_winpercent = mean(winpercent))
mean_fruity <- candy |> filter(fruity == 1) |> summarise(mean_winpercent = mean(winpercent))
mean_choc$mean_winpercent > mean_fruity$mean_winpercent

choc.inds <- as.logical(candy$chocolate)
choc.candy <- candy[choc.inds, ]
choc.win <- choc.candy$winpercent
choc.mean <- mean(choc.win)

fruit.inds <- as.logical(candy$fruity)
fruit.candy <- candy[fruit.inds, ]
fruit.win <- fruit.candy$winpercent
fruit.mean <- mean(fruit.win)

choc.mean > fruit.mean
```

> Q12. Is this difference statistically significant?

```{r}
t.test(candy$winpercent[candy$chocolate == 1], candy$winpercent[candy$fruity == 1])
```
Yes, they are statistically significantly different.

> Q13. What are the five least liked candy types in this set?

```{r}
candy[order(candy$winpercent), ][1:5, ]
```

> Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
head(candy[order(-candy$winpercent), ], 5)
```

> Q15. Make a first barplot of candy ranking based on winpercent values.

```{r}
ggplot(candy) + 
  aes(winpercent, rownames(candy)) +
  geom_col(fill = "lightgreen") +
  labs(title = "Candy Ranking Based on Winpercent Values", x = "Winpercent", y = "Candy Type") +
  theme_minimal(base_size = 12) +
  theme(axis.text.y = element_text(size = 7))
```
> Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy), winpercent)) +
  geom_col(fill = "lightgreen") +
  labs(title = "Candy Ranking Based on Winpercent Values", x = "Candy Type", y = "Winpercent") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Color mapping of histograms

```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"

ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols) 
```

> Q17. What is the worst ranked chocolate candy?

```{r}
candy_chocolate <- subset(candy, candy$chocolate == 1)
candy_chocolate[order(candy_chocolate$winpercent), ][1, ]
```

> Q18. What is the best ranked fruity candy?

```{r}
candy_fruity <- candy |> filter(fruity == 1) |> arrange(desc(winpercent)) |> slice(1)
candy_fruity
```

## Price percents 

```{r}
library(ggrepel)

# How about a plot of price vs win
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 15)
```

> Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

```{r}
candy_bang_for_buck <- candy %>%
  mutate(bang_for_buck = winpercent / pricepercent) %>%
  arrange(desc(bang_for_buck)) %>%
  slice(1)
candy_bang_for_buck
```

> Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

```{r}
candy_expensive <- candy %>%
  arrange(desc(pricepercent)) %>%
  slice(1:5)
candy_expensive
least_popular_expensive <- candy_expensive %>%
  arrange(winpercent) %>%
  slice(1)
least_popular_expensive # Nik L Nip
```

## Optional

> Q21. Make a barplot again with geom_col() this time using pricepercent and then improve this step by step, first ordering the x-axis by value and finally making a so called “dot chat” or “lollipop” chart by swapping geom_col() for geom_point() + geom_segment().

```{r}
ggplot(candy) +
  aes(pricepercent, reorder(rownames(candy), pricepercent)) +
  geom_col(fill = "lightcoral") +
  labs(title = "Candy Ranking Based on Pricepercent Values", x = "Pricepercent", y = "Candy Type") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
Lollipop Chart
```{r}
# Make a lollipop chart of pricepercent
ggplot(candy) +
  aes(pricepercent, reorder(rownames(candy), pricepercent)) +
  geom_segment(aes(yend = reorder(rownames(candy), pricepercent), 
                   xend = 0), col="gray40") +
    geom_point()
```

## Correlation Structures

```{r}
cor(candy)

library(corrplot)

cij <- cor(candy)
corrplot(cij)
```

> Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

Fruity and chocolate, bar and caramel, pluribus and bar, fruity and bar, etc.

> Q23. Similarly, what two variables are most positively correlated?

Chocolate and bar, bar and nougat, etc.

## PCA

Apply PCA using the prcom() function to our candy dataset remembering to set the scale=TRUE argument.

```{r}
pca <- prcomp(candy, scale = TRUE)
summary(pca)
```
Now we can plot our main PCA score plot of PC1 vs PC2.
```{r}
plot(pca$x[,1], pca$x[,2])
```
We can change the plotting character and add some color:
```{r}
plot(pca$x[,1:2], col=my_cols, pch=16)
```

We can make a much nicer plot with the ggplot2 package but it is important to note that ggplot works best when you supply an input data.frame that includes a separate column for each of the aesthetics you would like displayed in your final plot. To accomplish this we make a new data.frame here that contains our PCA results with all the rest of our candy data. We will then use this for making plots below

```{r}
# Make a new data-frame with our PCA results and candy data
my_data <- cbind(candy, pca$x[,1:3])

p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)

p

p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```


library(plotly)
# ggplotly() turns a ggplot into an interactive plotly plot
ggplotly(p)

Let’s finish by taking a quick look at PCA our loadings. Do these make sense to you? Notice the opposite effects of chocolate and fruity and the similar effects of chocolate and bar (i.e. we already know they are correlated).
```{r}
ggplot(as.data.frame(pca$rotation)) +
  aes(x=rownames(as.data.frame(pca$rotation)), y=PC1) +
  geom_col() +
  coord_flip() +
  labs(title="PCA Loadings for PC1",
       x="Variables",
       y="Loading Value") +
  theme_minimal(base_size = 12)
```
> Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

```{r}
# Variables picked up strongly in the positive direction for PC1
pca$rotation[which(pca$rotation[,1] > 0), 1]
```
Not really, thought fruity should be go along with chewy.
